\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-01-12 doc/publications/plos16/plos16.tex
consistent capitalization for package names plos16.tex

\newcommand{\cursedforest}{\textsc{CursedForest}\xspace}
\newcommand{\ranger}{\textsc{Ranger}\xspace}
\newcommand{\mtry}{\texttt{mtry}\xspace}
\newcommand{\ntree}{\texttt{ntree}\xspace}

@@

The most popular random forest software packages are R-based, although current implementations do not scale well with
increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}-1$), and the \textsc{randomForest} implementation in R lending itself to parallelization to grow multiple
trees simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide''
genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015} written in Go,
achieves fast run times by effective use of the CPU cache, optimizing for different classes of features and efficiently
multi-threading.  Similarly, \ranger~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is
specifically optimized for dealing with large data sets.

@@

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R notation from the \textsc{randomForest}
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in
  \textsc{randomForest} by setting \texttt{nodesize}, the minimum size of
  terminal nodes. It is controlled in Spark ML by setting
  \texttt{maxDepth}, the maximum depth of each tree in the
  forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets
    \texttt{maxDepth} to 4 in their classification example. This may
    give an estimate with a higher bias and, as such, may be a poor
    choice. See the discussion in \cite{Dietterich.2002} where it is
    shown that bagging may reduce bias as well as variance but gives a
    better result for low bias learners. \cite{Dietterich.2002} notes
    that if the boostrap replicate approximation were correct, then
    bagging would reduce variance without changing bias.} In the TCGA
  example the RandomForests code produced trees with depths of 30+.
\end{itemize}

@@

\multirow{5}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.04  & 1 min 50 seconds                        \\
                                                & Spark ML  &  0.05           &           7 min 44 seconds         \\
                                                & randomForest (R)         & 0.05       & 24 min           \\
                                                & ranger (R)      & 0.06       & 1 min 10 seconds       \\
                                                & ranger (C++)     & 0.06       & 7 min            \\
                                                & H2O           & ??       & ??         \\

@@

\multirow{5}{*}{phase3\_chr22: 2,504 x 1,103,548}   & \cursedforest\ & 0.06 & 4 min 2 seconds \\
                                                    & Spark ML  & 0.06 & 37 min 19 seconds \\
                                                    & randomForest (R)        & NA         & NA                \\
                                                    & ranger (R)    & 0.06       & 4 min 30 seconds     \\
                                                    & ranger (C++)     &     ??   & ??        \\
                                                    & H2O           & 0.05       & 16 hours         \\

@@

\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.04  & 19 min 19 seconds         \\
                                                    & Spark ML  & 0.04       & 8 hour 29 min        \\
                                                    & randomForest (R)        & NA         & NA                \\
                                                    & ranger (R)      &  0.05       & 28 mins         \\
                                                    & ranger (C++)     & ??       & ??            \\
                                                    & H2O           & NA       & NA         \\
\hline

\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.02  & 56 min 43 seconds             \\
                                                    & Spark ML &    NA         &   NA               \\
                                                    & randomForest (R)        & NA         & NA                \\
                                                    & ranger (R)      & 0.03      &    3 hour 58 min     \\
                                                    & ranger (C++)     & ??       & ??            \\
                                                    & H2O           & NA       & NA         \\
\hline

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & 0.01  & 3hour 16min \\
                                                    & Spark ML & NA & NA  \\
                                                    & randomForest (R)        & NA         & NA                \\
                                                    & ranger (R)       &        NA     &        NA    \\
                                                    & ranger (C++)       &        NA     &        NA    \\
                                                    & H2O           & NA       & NA         \\
\hline

@@

In order to asses these characteristics we ran \cursedforest\ classification on synthetics datasets with varying numbers
of variables (features) and samples, similar to the dataset used in \cite{Wright.and.Ziegle.2016} to evaluate
\textsc{Ranger}, allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity
of random forests by using a range of \mtry values.

>>>>

\newcommand{\cursedforest}{\textsc{CursedForest}\xspace}
\newcommand{\ranger}{\textsc{Ranger}\xspace}
\newcommand{\randomforest}{\textsc{randomForest}\xspace}
\newcommand{\mtry}{\texttt{mtry}\xspace}
\newcommand{\ntree}{\texttt{ntree}\xspace}

@@

The most popular random forest software packages are R-based, although current implementations do not scale well with
increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}-1$), and the \randomforest implementation in R lending itself to parallelization to grow multiple
trees simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide''
genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015} written in Go,
achieves fast run times by effective use of the CPU cache, optimizing for different classes of features and efficiently
multi-threading.  Similarly, \ranger~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is
specifically optimized for dealing with large data sets.

@@

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R notation from the \randomforest
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in
  \randomforest by setting \texttt{nodesize}, the minimum size of
  terminal nodes. It is controlled in Spark ML by setting
  \texttt{maxDepth}, the maximum depth of each tree in the
  forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets
    \texttt{maxDepth} to 4 in their classification example. This may
    give an estimate with a higher bias and, as such, may be a poor
    choice. See the discussion in \cite{Dietterich.2002} where it is
    shown that bagging may reduce bias as well as variance but gives a
    better result for low bias learners. \cite{Dietterich.2002} notes
    that if the boostrap replicate approximation were correct, then
    bagging would reduce variance without changing bias.} In the TCGA
  example the RandomForests code produced trees with depths of 30+.
\end{itemize}

@@

\multirow{5}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.04  & 1 min 50 seconds                        \\
                                                & Spark ML  &  0.05           &           7 min 44 seconds         \\
                                                & \randomforest (R)         & 0.05       & 24 min           \\
                                                & \ranger (R)      & 0.06       & 1 min 10 seconds       \\
                                                & \ranger (C++)     & 0.06       & 7 min            \\
                                                & H2O           & ??       & ??         \\

@@

\multirow{5}{*}{phase3\_chr22: 2,504 x 1,103,548}   & \cursedforest\ & 0.06 & 4 min 2 seconds \\
                                                    & Spark ML  & 0.06 & 37 min 19 seconds \\
                                                    & \randomforest (R)        & NA         & NA                \\
                                                    & \ranger (R)    & 0.06       & 4 min 30 seconds     \\
                                                    & \ranger (C++)     &     ??   & ??        \\
                                                    & H2O           & 0.05       & 16 hours         \\

@@

\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.04  & 19 min 19 seconds         \\
                                                    & Spark ML  & 0.04       & 8 hour 29 min        \\
                                                    & \randomforest (R)        & NA         & NA                \\
                                                    & \ranger (R)      &  0.05       & 28 mins         \\
                                                    & \ranger (C++)     & ??       & ??            \\
                                                    & H2O           & NA       & NA         \\
\hline

\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.02  & 56 min 43 seconds             \\
                                                    & Spark ML &    NA         &   NA               \\
                                                    & \randomforest (R)        & NA         & NA                \\
                                                    & \ranger (R)      & 0.03      &    3 hour 58 min     \\
                                                    & \ranger (C++)     & ??       & ??            \\
                                                    & H2O           & NA       & NA         \\
\hline

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & 0.01  & 3hour 16min \\
                                                    & Spark ML & NA & NA  \\
                                                    & \randomforest (R)        & NA         & NA                \\
                                                    & \ranger (R)       &        NA     &        NA    \\
                                                    & \ranger (C++)       &        NA     &        NA    \\
                                                    & H2O           & NA       & NA         \\
\hline

@@

In order to asses these characteristics we ran \cursedforest\ classification on synthetics datasets with varying numbers
of variables (features) and samples, similar to the dataset used in \cite{Wright.and.Ziegle.2016} to evaluate
\ranger, allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity
of random forests by using a range of \mtry values.
